---
title: "Covid-19 in a United States County"
author: "Udayan Roy"
date: "`r Sys.Date()`"
output:
  html_document: default
params:
  fips: "17043"
  state_long: "Illinois"
  state_short: "IL"
  county: "DuPage"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data

Here I use the data available at <https://github.com/nytimes/covid-19-data> to highlight a few aspects of the Covid-19 pandemic in `r params$county` county, `r params$state_long`. I show the *R* code that I have used.

I begin by making the *tidyverse* package available for use:

```{r message=FALSE, warning=FALSE}
library(tidyverse)
```

Next, I download the data provided by The New York Times for all US counties and prepare a smaller data frame for `r params$county` county, `r params$state_short`, by specifying its [FIPS code](https://en.wikipedia.org/wiki/FIPS_county_code), which happens to be `r params$fips`:

```{r message=FALSE, warning=FALSE}
corona.us.counties <- read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv")
mycounty.mystate <- filter(corona.us.counties, fips == params$fips)
head(mycounty.mystate)
```

The crucial variables are *cases* and *deaths*, representing cumulative counts. Note that the data are arranged chronologically and begin on `r strftime(mycounty.mystate$date[[1]], "%A, %B %d, %Y")`, the day the first case was recorded in `r params$county` county, `r params$state_long`.

## Graphs of Cumulative Covid-19 Cases and Deaths

### Cumulative Covid-19 Cases

As an example of the use of the `xts` and `dygraphs` packages, I present an interactive graph of the cumulative number of Covid-19 cases. The `xts` package is widely used to work with time series data. The `dygraphs` package creates, inter alia, interactive graphs when it is fed an `xts` data object.
```{r message=FALSE, warning=FALSE}
library(xts)
library(dygraphs)
mycounty.mystate.coredata <- mycounty.mystate %>%
  select(cases, deaths) 
mycounty.mystate.index <- as.Date(mycounty.mystate$date, "%m/%d/%Y")

mycounty.mystate.xts <- xts(mycounty.mystate.coredata, order.by= mycounty.mystate.index)

dygraph(mycounty.mystate.xts$cases, main = "Cumulative Cases", width = 500, height = 300) %>%
  dyRangeSelector() %>%
  dyHighlight(highlightCircleSize = 5,
              highlightSeriesBackgroundAlpha = 0.2,
              hideOnMouseOut = FALSE)
# To show all the variables in the `xts` object, delete `$cases`.
```

Note, again, that this graph is __interactive__! If you glide your cursor over the graph, you should see an ever-changing label giving the cumulative cases of Covid-19 for the relevant day. You should also be able to drag the sliders on the graph's horizontal axis to choose the beginning and end of the chart's time period.

The next graph shows the same data as the one above, but using a logarithmic scale (with one unit of height along the vertical scale representing a doubling of the plotted variable). Moreover, this is a static -- that is, non-interactive -- graph. 

For my static graphs, I use the *ggplot2* package, which is part of the *tidyverse* package that I have already made ready for use.

```{r message=FALSE, warning=FALSE}
ggplot(data = mycounty.mystate) +    
  geom_point(mapping = aes(x = date, y = cases), color = "blue") +
  scale_y_continuous(trans = 'log2') +
  labs(x = "Date", y = "Cumulative Cases", title = "The Spread of the Virus", subtitle = "Logarithmic Scale")
```

### Cumulative Covid-19 Deaths

```{r message=FALSE, warning=FALSE}
ggplot(data = mycounty.mystate) +  
  geom_line(mapping = aes(x = date, y = deaths)) +
  labs(x = "Date", y = "Cumulative Deaths", title = "The Toll", subtitle = "Linear Scale")
```

And in logarithmic scale:

```{r message=FALSE, warning=FALSE}
ggplot(data = mycounty.mystate) +    
  geom_point(mapping = aes(x = date, y = deaths), color = "blue") +
  scale_y_continuous(trans = 'log2') +
  labs(x = "Date", y = "Cumulative Deaths", title = "The Toll", subtitle = "Logarithmic Scale")
```

And, having graphed the data for Covid-19 cases and deaths, it is not too much of a detour to look at the [Case Fatality Rate](https://en.wikipedia.org/wiki/Case_fatality_rate), which is deaths as a percent of cases:

```{r message=FALSE, warning=FALSE}
ggplot(data = mycounty.mystate) +  
  geom_line(mapping = aes(x = date, y = 100*(deaths/cases))) +
  labs(x = "Date", y = "Deaths as a percent of Cases", title = "Case Fatality Rate")
```

Note that this rate would depend heavily on the number of tests being done and on the criteria used to determine who gets tested. Moreover, this case fatality rate is _cumulative_ deaths as a percent of _cumulative_ cases. As time passes and the pandemic matures, day to day changes in these cumulative numbers will be relatively inconsequential. Consequently, the CFR, being a ratio of slow-changing numbers, will itself be slow to change.

## Daily Numbers of New Covid-19 Cases and Deaths 

The increase in the cumulative totals from one date to the next gives the increment for the second of the two dates. The *seven-day averages of the daily increases* are also calculated.

```{r message=FALSE, warning=FALSE}
mycounty.mystate <- mycounty.mystate %>%
  arrange(date) %>%   # This is not strictly necessary
  mutate(increase.in.cases = cases - lag(cases), 
         increase.in.deaths = deaths - lag(deaths)) %>%
  mutate(increase.in.cases.7days = (cases - lag(cases, 7))/7,
         increase.in.deaths.7days = (deaths - lag(deaths, 7))/7)
```

Now the *daily tallies* of new cases and deaths can be graphed, with the *seven-day averages* overlaid in blue:

### Daily Tally of New Covid-19 Cases and the Seven-Day Average

```{r message=FALSE, warning=FALSE}
ggplot(data = mycounty.mystate) +  
  geom_line(mapping = aes(x = date, y = increase.in.cases)) +
  geom_line(mapping = aes(x = date, y = increase.in.cases.7days), color = "blue", linetype = 1, size = 1.5) +
  labs(x = NULL, y = NULL, title = "The Daily Increase in Cases and its Seven-Day Average")
```

### Daily Tally of New Covid-19 Deaths and the Seven-Day Average

```{r message=FALSE, warning=FALSE}
ggplot(data = mycounty.mystate) +  
  geom_line(mapping = aes(x = date, y = increase.in.deaths)) +
  geom_line(mapping = aes(x = date, y = increase.in.deaths.7days), color = "blue", linetype = 1, size = 1.5) +
  labs(x = NULL, y = NULL, title = "The Daily Increase in Deaths and its Seven-Day Average")
```


## The Toughest Days So Far in `r params$county` County, `r params$state_short`

### These Were the Days with the Most New Cases

```{r message=FALSE, warning=FALSE}
tail(mycounty.mystate %>% select(date, increase.in.cases, increase.in.deaths) %>% arrange(increase.in.cases) %>% na.omit()) 
```

### These Were the Days with the Most New Deaths

```{r message=FALSE, warning=FALSE}
tail(mycounty.mystate %>% select(date, increase.in.cases, increase.in.deaths) %>% arrange(increase.in.deaths) %>% na.omit())
```

## The Last Two Weeks

```{r message=FALSE, warning=FALSE}
tail(mycounty.mystate %>% select(date, increase.in.cases, increase.in.deaths), 14)
```

### Last Two Weeks: Daily Tally of New Cases

```{r message=FALSE, warning=FALSE}
ggplot(data = tail(mycounty.mystate, 14)) +  
  geom_line(mapping = aes(x = date, y = increase.in.cases))
```

### Last Two Weeks: Daily Tally of New Deaths

```{r message=FALSE, warning=FALSE}
ggplot(data = tail(mycounty.mystate, 14)) +   
  geom_line(mapping = aes(x = date, y = increase.in.deaths))
```

## Conclusion

Needless to say, the code here can be used to present a similar profile for any other US county, by inserting the appropriate *fips* number for the county in the first of my code chunks. 

This essay is meant to help me remember the *R* commands I used in it. I am an amateur "data scientist" and I work on simple projects on occasion. As a result of the long gaps between my "projects", I tend to forget what I learn. 
